\sys: A Proof-of-Concept for Explainable Disease Diagnosis Using Large Language Models and Answer Set Programming

### 2025-12-30 paper/draft.tex
Fixes
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\usepackage{datetime}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{array,tabularx}
\usepackage{chngcntr}
\usepackage{afterpage}
\usepackage{ulem}
\usepackage{hyperref}
\usepackage{dirtytalk}
\usepackage{algorithm2e}
\usepackage{enumitem,amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{formal-grammar}
\usepackage{varwidth}
\usepackage{fmtcount}
\usepackage{tikz}
\usepackage{array}
\usepackage[framemethod=tikz]{mdframed} % Allows defining custom boxed/framed environments
\usepackage{cleveref}
\usepackage{microtype}
\usepackage{dirtytalk}
\usepackage{booktabs}

@@

@@

\mdfdefinestyle{warning}{
	topline=false, bottomline=false,
	leftline=false, rightline=false,
	nobreak,
	singleextra={%
		\draw(P-|O)++(-0.5em,0)node(tmp1){};
		\draw(P-|O)++(0.5em,0)node(tmp2){};
		\fill[black,rotate around={45:(P-|O)}](tmp1)rectangle(tmp2);
		\node at(P-|O){\color{white}\scriptsize\bf ?};
		\draw[very thick](P-|O)++(0,-1em)--(O);%--(O-|P);
	}
}

@@

\title[\sys: Automatic Disease Diagnosis Using LLMs \& ASP]{\sys: Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming}

\author[Ioanna Gemou]{Ioanna Gemou}
\author[Evangelos Lamprou]{Evangelos Lamprou}
\thanks{Originally submitted as part of a research competition on 1st June 2023.}

@@

\maketitle
\vspace{-1.5em} % adjust as needed
{\small\itshape \centering Originally submitted as part of a university research competition on 1st June 2023.\par}
\vspace{0.5em} % optional: add space before body

\begin{abstract}
Accurate disease prediction is vital for timely intervention, 
effective treatment, and reducing medical complications. 
While symbolic AI has been applied in healthcare, 
its adoption remains limited due to the complexity 
of required knowledge bases.
This work introduces \sys, a framework that combines Large Language Models (LLMs) 
with Answer Set Programming (ASP) to overcome this barrier. 
LLMs extract medical knowledge from text and translate it into ASP code, 
while ASP enables structured, explainable reasoning. 
This integration yields a robust, interpretable prediction framework 
that leverages the strengths of both paradigms. 
\sys shows strong performance on real patient data.
% \sys accurately predicts diseases based on symptoms derived from medical literature.
\end{abstract}

\section{Introduction}
Getting an accurate diagnosis quickly can make all the difference in patient outcomes, 
but building automated systems that practitioners can actually trust is still a major challenge. 
Symbolic AI, and in particular Answer Set Programming (ASP), provides clear logical reasoning \cite{Brewka2011,Gelfond_Kahl_2014}. 
However, its adoption has been limited because constructing and maintaining large, 
domain-specific knowledge bases is very labor-intensive.
On the other hand, Large language models (LLMs) achieve strong results on tasks
such as medical question answering and clinical note summarization \cite{Singhal2022LargeLM,Singhal2023,Zheng2024LLMmed}, 
but their tendency to produce inconsistent or opaque outputs makes them risky for clinical use \cite{aljohani2025comprehensivesurveytrustworthinesslarge, rudin2019stopexplainingblackbox}.

In clinical decision support, the real goal is to get the best of both worlds: 
systems that reason accurately while still being easy for practitioners to trust and use. 
\sys takes a step in that direction by combining the strengths of LLMs and ASP. 
Instead of requiring experts to hand-craft large knowledge bases, 
\sys automatically converts medical literature into ASP rules. 
These rules are then enriched with patient symptoms and clinical indicators, 
and passed through an ASP solver to produce diagnostic suggestions together with logical justifications.
On public patient records, \sys reaches 95–100\% predictive accuracy on selected diseases, 
while giving transparent explanations for each diagnosis.

@@

\paragraph{\textbf{Large Language Models}}
LLMs \cite{zhao2023survey} are advanced computational systems 
designed to understand and generate human language. 
Developed using cutting-edge machine learning techniques, 
these models can produce text that closely mirrors human linguistic patterns.
Modern architectures such as \textit{GPT-5} \cite{openai2023gpt4, gpt5}, 
are trained on massive corpora spanning diverse domains, 
enabling them to acquire broad, cross-disciplinary knowledge. 
LLMs use deep learning algorithms \cite{Sarker2021} 
and specifically transformer networks \cite{Dosovitskiy2020} 
to analyze and understand the relationships between words, phrases, and sentences. 
By capturing patterns and structures in the training data, 
they can generate coherent and relevant responses based on given prompts.
The term “large” refers to the billions of parameters 
that allow these models to generate context-aware responses. 
The basic premise of an LLM is its ability to predict 
the next word (referred to as a token) based on the text it has observed so far, 
according to the context.

@@

LLMs have demonstrated strong performance across a variety of natural language tasks, 
including question answering \cite{Brown2020}, summarization \cite{Zhang2020}, 
dialogue generation \cite{Thoppilan2022lamda}, and code synthesis \cite{Chen2021}. 
In specialized domains such as biomedicine, domain-adapted LLMs have been employed 
for clinical decision support \cite{Singhal2023} and 
biomedical knowledge retrieval \cite{Luo2022}.
Among their capabilities, LLMs excel at transforming text from one representation to another, 
making them powerful tools for a wide range of natural language processing tasks. 
This characteristic is particurarly relevant to the present work, 
which focuses on the effective encoding of medical literature. \\

@@

Answer Set Programming (ASP) \cite{Eiter2009} is a declarative problem-solving paradigm 
rooted in logic programming and non-monotonic reasoning \cite{Brewka2011}. 
The formal semantics of stable models and the foundational ASP language 
were first introduced by Gelfond and Lifschitz \cite{Gelfond2000, gel88}.
Programming with this approach is done in a family of languages 
sometimes referred to as \textit{AnsProlog} \cite{Gelfond2002}.

@@

In traditional programming, transitioning from a problem to a solution 
requires the programmer's thorough understanding of the given problem, 
followed by the creation of a program that will produce the correct output 
when provided with an instance of the problem, which will then be interpreted as the solution. 
In ASP, the process of deriving solutions from a problem specification 
typically follows a sequence of well-defined steps \cite{Gebser2013}, 
as illustrated in \cref{fig:asp-solving}. 
First, the problem is modeled using the ASP language. 
Next, a grounder (e.g., gringo \cite{Gebser2014}) transforms the program into a set of basic rules and facts. 
Finally, a solver (e.g., clasp \cite{Holldobler2014}) computes the stable models (answer sets), 
which correspond to the solutions of the original problem.

@@

        \draw [arrow] (problem) -- node[right] {\textbf{Modelling}} (program);
        \draw [arrow] (stable-models) -- node[left] {Interpreting} (solution);

@@

Several solvers for ASP have been developed, such as DLV \cite{Xia2020}. 
This work uses the ASP system clingo, 
which integrates the gringo grounder \cite{Gebser2014} with the clasp solver \cite{Holldobler2014} into a single system. 
Clingo also offers a powerful Python API, enabling the solver's integration into other applications.

@@

\Cref{fig:architecture} illustrates the \sys architecture, 
which transforms unstructured medical text into a structured knowledge base \cite{ré2014feature}. 
The process begins with the collection of medical literature relevant to the diseases or conditions of interest. 
A carefully designed prompt is then used to guide an LLM in translating this literature into ASP code, 
which is incorporated into a growing knowledge base. 
This knowledge base is further enriched with patient-specific information such as symptoms and examination results. 
An ASP solver subsequently reasons over the combined knowledge base and produces a diagnosis
--or a set of candidate diagnoses--in the form of answer sets.

@@

\paragraph{\textbf{Prompt Construction}}
While large language models exhibit strong instruction-following capabilities, 
their reliability in high-stakes domains remains limited \cite{bowman2023things}. 
To improve robustness, a wide range of prompting strategies have been developed 
and refined over the past several years \cite{liu2021pretrain, promptingguidePromptEngineering}. 
In \sys, carefully designed prompts are employed to guide the model in translating medical literature 
into ASP code suitable for execution by a solver (\cref{fig:bad-prompt}, \cref{fig:good-prompt}).
\footnote{The example of a poorly constructed prompt is included for educational purposes, to highlight common pitfalls in prompt design.}

@@

The paragraph above lists common symptoms of pneumonia.
Write a Clingo script that diagnoses pneumonia based on these symptoms.
\end{prompt}

@@

The paragraph above lists common symptoms of pneumonia.
Write a Clingo script that diagnoses pneumonia based on these symptoms.

In the diagnosis rule, use a structure like:

@@

Include alternative diagnoses that share overlapping symptoms with pneumonia.
Add rules that link one symptom to another (e.g., symptom propagation or dependency).

@@

The knowledge base is initialized with a set of symptoms and medical indicators represented as:
\begin{align}
    symptom(s). & \; s \in \{ \text{cough, chest pain, rash}, \dots \} \\
    indicator(i). & \; i \in \{ \text{low MCV, high TIBC}, \dots \}
\end{align}

A patient exhibiting a specific symptom or testing positive for a particular indicator is represented as:
\begin{equation}
    has(x). \; x \in S \cup I
\end{equation}

@@

To ensure that the solver identifies at least one diagnosis, 
the following integrity constraint is imposed:
\begin{equation}
    \bot \longleftarrow \sim diagnosis(\_)
\end{equation}

Finally, the solver minimizes assumptions by penalizing added symptoms, 
favoring diagnoses based on known patient information:
\begin{equation}
    \#minimize \{ 1, S : add(S) \}
\end{equation}

@@

Symbolic Artifical Intelligence offers a significant advantage 
over subsymbolic methods in terms of explainability. 
In symbolic systems, the relationships between facts and conclusions are explicitly defined, 
enabling transparent reasoning. However, ASP solvers such as \textit{Clingo} do not, 
at the time of writing,\footnote{At the time of writing this paper.} 
provide intuitive traces of their reasoning processes. 
Prior work has explored methods for associating logical propositions 
with their justifications \cite{cabalar2014causal}. For example, 
the logic program shown below is accompanied by a causal graph 
in \cref{fig:causal-g}, which illustrates the connections 
between rules and their inferred outcomes. 
This structure supports the interpretability of conclusions derived by a symbolic logic solver.
% \begin{equation}
% \begin{aligned}[b]
%     l :& punish \longleftarrow drive, drunk \\
%     m :& punish \longleftarrow resist \\
%     e :& prison \longleftarrow punish \\
%     d :& drive \\
%     k :& drunk \\
%     r :& resist \\
% \end{aligned}
% \label{eq:example-cg}
% \end{equation}
% % https://docs.google.com/drawings/d/1Ing4GFH2kbniiHI5rbXzw-xcAUQAxWAJpgmg4tz1fzc/edit?usp=sharing
% \begin{figure}
%     \centering
%     \includegraphics{assets/causal_g.pdf}
%     \caption{Causal graph of logic program (\cref{eq:example-cg}).}
%     \label{fig:causal-g}
% \end{figure}
\begin{figure}[t]
    \centering
    \begin{minipage}{0.48\textwidth}
    \begin{equation}
    \begin{aligned}[b]
        l :& punish \longleftarrow drive, drunk \\
        m :& punish \longleftarrow resist \\
        e :& prison \longleftarrow punish \\
        d :& drive \\
        k :& drunk \\
        r :& resist \\
    \end{aligned}
    \label{eq:example-cg}
    \end{equation}
    \end{minipage}
    \hfill
    \begin{minipage}{0.45\textwidth}
    \centering
    \includegraphics[width=.5\textwidth]{assets/causal_g.pdf}
    \caption{Causal graph of logic program (\cref{eq:example-cg}).}
    \label{fig:causal-g}
    \end{minipage}
\end{figure}

The tool \textit{xclingo}~\cite{Cabalar_2020} 
enables the visualization of justifications for results produced by the \textit{Clingo} solver. 
It does so by incorporating a \src{trace} that corresponds to the rules of the logic program. 
Providing clear explanations of how the framework reaches specific conclusions helps users identify 
potential errors in the construction of the knowledge base and offers transparency for the final outcome.

@@

Evaluation was conducted using publicly available patient data.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}
Each patient's list of symptoms serves as input.  
The knowledge base—comprising ASP rules automatically derived  
from medical literature—infers potential diagnoses.  
A prediction is considered correct when the disease inferred  
by the ASP solver matches the actual diagnosis listed in the dataset.

\begin{table}[h]
    \centering
    \caption{Accuracy of the framework on selected diseases. 
    The size column indicates the number of logical terms in the disease-specific ASP program.}
    \begin{tabular}{lcc}
    \toprule
    Diseases & Size & Accuracy \\
    \midrule
    Chicken Pox    &  66  & 95\% \\
    Pneumonia      &  75  & 100\% \\
    Common Cold    &  44  & 100\% \\
    \bottomrule
    \end{tabular}
\end{table}

@@

Future work will focus on grounding ASP rules in reliable sources 
through retrieval-augmented generation, fine-tuning LLMs for clinical reasoning, 
and scaling the framework to cover a wider range of diseases.
\bibliographystyle{unsrt}
\bibliography{bib}

>>>>

\usepackage{datetime}
\usepackage{wrapfig}
\usepackage{enumerate}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{array,tabularx}
\usepackage{chngcntr}
\usepackage{afterpage}
\usepackage{ulem}
\usepackage{hyperref}
\usepackage{dirtytalk}
\usepackage{algorithm2e}
\usepackage{enumitem,amssymb}
\usepackage{pifont}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{formal-grammar}
\usepackage{varwidth}
\usepackage{fmtcount}
\usepackage{tikz}
\usepackage[framemethod=tikz]{mdframed} % Allows defining custom boxed/framed environments
\usepackage{cleveref}
\usepackage{microtype}
\usepackage{booktabs}

@@

\newcommand{\eg}{{e.g.}, }
\newcommand{\ie}{{i.e.}, }
\newcommand{\etc}{{etc.}\xspace}
\newcommand{\vs}{{vs.} }
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\sx}[1]{(\cref{#1})}

@@

\mdfdefinestyle{warning}{
	topline=false, bottomline=false,
	leftline=false, rightline=false,
	nobreak,
	singleextra={%
		\draw(P-|O)++(-0.5em,0)node(tmp1){};
		\draw(P-|O)++(0.5em,0)node(tmp2){};
		\fill[black,rotate around={45:(P-|O)}](tmp1)rectangle(tmp2);
		\node at(P-|O){\color{white}\scriptsize\bf !};
		\draw[very thick](P-|O)++(0,-1em)--(O);%--(O-|P);
	}
}

@@

\title[\sys: A PoC for Explainable Disease Diagnosis Using LLMs \& ASP]{\sys: A Proof-of-Concept for Explainable Disease Diagnosis Using Large Language Models and Answer Set Programming}

\author[Ioanna Gemou]{Ioanna Gemou}
\author[Evangelos Lamprou]{Evangelos Lamprou}
\thanks{Originally submitted as part of a university research competition on 1st June 2023.}

@@

\maketitle

\begin{abstract}
Accurate disease prediction is vital for timely intervention, 
effective treatment, and reducing medical complications. 
While symbolic AI has been applied in healthcare, 
its adoption remains limited due to the complexity 
of required knowledge bases.
This work introduces \sys, a framework that combines Large Language Models (LLMs) 
with Answer Set Programming (ASP) to overcome this barrier. 
LLMs extract medical knowledge from text and translate it into ASP code, 
while ASP enables structured, explainable reasoning. 
This integration yields a robust, interpretable prediction framework 
that leverages the strengths of both paradigms. 
\sys shows strong performance on public datasets.
% \sys accurately predicts diseases based on symptoms derived from medical literature.
\end{abstract}

\section{Introduction}
Getting an accurate diagnosis quickly can make all the difference in patient outcomes, 
but building automated systems that practitioners can actually trust is still a major challenge. 
Symbolic AI, and in particular Answer Set Programming (ASP), provides clear logical reasoning \cite{Brewka2011,Gelfond_Kahl_2014}. 
However, its adoption has been limited because constructing and maintaining large, 
domain-specific knowledge bases is very labor-intensive.
On the other hand, Large Language Models (LLMs) achieve strong results on tasks
such as medical question answering and clinical note summarization \cite{Singhal2022LargeLM,Singhal2023,Zheng2024LLMmed}, 
but their tendency to produce inconsistent or opaque outputs makes them risky for clinical use~\cite{aljohani2025comprehensivesurveytrustworthinesslarge, rudin2019stopexplainingblackbox}.

In clinical decision support, the real goal is to get the best of both worlds: 
systems that reason accurately while still being easy for practitioners to trust and use. 
\sys takes a step in that direction by combining the strengths of LLMs and ASP. 
Instead of requiring experts to hand-craft large knowledge bases, 
\sys automatically converts medical literature into ASP rules. 
These rules are then enriched with patient symptoms and clinical indicators, 
and passed through an ASP solver to produce diagnostic suggestions together with logical justifications.
On public datasets, \sys reaches 95–100\% predictive accuracy on selected diseases, 
while giving transparent explanations for each diagnosis.

@@

\paragraph{\textbf{Large Language Models}}
LLMs \cite{zhao2023survey} are advanced computational systems 
designed to understand and generate human language. 
Developed using cutting-edge machine learning techniques, 
these models can produce text that closely mirrors human linguistic patterns.
Modern architectures such as \textit{GPT-5}~\cite{gpt5}
are trained on massive corpora spanning diverse domains, 
enabling them to acquire broad, cross-disciplinary knowledge. 
LLMs use deep learning algorithms \cite{Sarker2021} 
and specifically transformer networks \cite{Vaswani2017} 
to analyze and understand the relationships between words, phrases, and sentences. 
By capturing patterns and structures in the training data, 
they can generate coherent and relevant responses based on given prompts.
The term ``large" often refers to models with billions of parameters 
that allow these models to generate context-aware responses. 
The basic premise of an LLM is its ability to predict 
the next word (referred to as a token) based on the text it has observed so far, 
according to the context.

@@

LLMs have demonstrated strong performance across a variety of natural language tasks, 
including question answering \cite{Brown2020}, summarization \cite{Zhang2020}, 
dialogue generation \cite{Thoppilan2022lamda}, and code synthesis \cite{Chen2021}. 
In specialized domains such as biomedicine, domain-adapted LLMs have been employed 
for clinical decision support \cite{Singhal2023} and 
biomedical knowledge retrieval \cite{Luo2022}.
Among their capabilities, LLMs excel at transforming text from one representation to another, 
making them powerful tools for a wide range of natural language processing tasks. 
This characteristic is particularly relevant to the present work, 
which focuses on the effective encoding of medical literature. \\

@@

Answer Set Programming (ASP) \cite{Eiter2009} is a declarative problem-solving paradigm 
rooted in logic programming and non-monotonic reasoning \cite{Brewka2011}. 
The formal semantics of stable models and the foundational ASP language 
were first introduced by Gelfond and Lifschitz \cite{Gelfond2000, gel88}.
These are implemented in a family of languages commonly known as \textit{AnsProlog} \cite{Gelfond2002}.

@@

In traditional programming, transitioning from a problem to a solution 
requires the programmer's thorough understanding of the given problem, 
followed by the creation of a program that will produce the correct output 
when provided with an instance of the problem, which will then be interpreted as the solution. 
In ASP, the process of deriving solutions from a problem specification 
typically follows a sequence of well-defined steps \cite{Gebser2013}, 
as illustrated in \cref{fig:asp-solving}. 
First, the problem is modeled using the ASP language. 
Next, a grounder (\eg gringo \cite{Gebser2014}) transforms the program into a set of basic rules and facts. 
Finally, a solver (\eg clasp \cite{Holldobler2014}) computes the stable models (answer sets), 
which correspond to the solutions of the original problem.

@@

        \draw [arrow] (problem) -- node[right] {\textbf{Modeling}} (program);
        \draw [arrow] (stable-models) -- node[left] {Interpreting} (solution);

@@

Several solvers for ASP have been developed, such as DLV \cite{Xia2020}. 
This work uses the ASP system \src{clingo}, 
which integrates the gringo grounder \cite{Gebser2014} with the clasp solver \cite{Holldobler2014} into a single system. 
The Clingo system also offers a powerful Python API, enabling integration into other applications.

@@

\Cref{fig:architecture} illustrates the \sys architecture, 
which transforms unstructured medical text into a structured knowledge base \cite{ré2014feature}. 
The process begins with the collection of medical literature relevant to the diseases or conditions of interest. 
A carefully designed prompt is then used to guide an LLM in translating this literature into ASP code, 
which is incorporated into a growing knowledge base. 
This knowledge base is further enriched with patient-specific information such as symptoms and examination results. 
An ASP solver subsequently reasons over the combined knowledge base and produces a diagnosis, or a set of plausible diagnoses.

@@

\paragraph{\textbf{Prompt Construction}}
While Large Language Models exhibit strong instruction-following capabilities, 
their reliability in high-stakes domains remains limited \cite{bowman2023things}. 
To improve robustness, a wide range of prompting strategies have been developed 
and refined over the past several years \cite{liu2021pretrain, promptingguidePromptEngineering}. 
In \sys, carefully designed prompts are employed to guide the model in translating medical literature 
into ASP code suitable for execution by a solver (\cref{fig:bad-prompt}, \cref{fig:good-prompt}).
\footnote{The example of a poorly constructed prompt is included for educational purposes, to highlight common pitfalls in prompt design.}

@@

The paragraph above lists common symptoms of pneumonia.
Write a clingo script that diagnoses pneumonia based on these symptoms.
\end{prompt}

@@

The paragraph above lists common symptoms of pneumonia.
Write a clingo script that diagnoses pneumonia based on these symptoms.
In the diagnosis rule, use a structure like:

@@

Include alternative diagnoses that share overlapping symptoms with pneumonia.
Add rules that link one symptom to another (\eg symptom propagation or dependency).

@@

The knowledge base is initialized with a set of symptoms and medical indicators represented as:
\begin{align}
    symptom(s). & \; s \in \{ \text{cough, chest pain, rash}, \dots \} \\
    indicator(i). & \; i \in \{ \text{low MCV, high TIBC}, \dots \}
\end{align}
Let $S$ denote the set of symptoms and $I$ the set of indicators.

A patient exhibiting a specific symptom or testing positive for an indicator is represented as:
\begin{equation}
    has(x). \; x \in S \cup I
\end{equation}

@@

The following integrity constraint imposes at least one diagnosis:
\begin{equation}
    \bot \longleftarrow \sim diagnosis(\_)
\end{equation}

Finally, the solver minimizes assumptions by penalizing added symptoms, 
favoring diagnoses based on known patient information:
\begin{equation}
    \#minimize \{ 1, S : add(symptom(S)) \}
\end{equation}

@@

Symbolic Artificial Intelligence offers a significant advantage 
over subsymbolic methods in terms of explainability. 
In symbolic systems, the relationships between facts and conclusions are explicitly defined, 
enabling transparent reasoning. Prior work has explored methods for associating logical propositions 
with their justifications \cite{cabalar2014causal}. For example, 
the logic program shown below is accompanied by a causal graph 
in \cref{fig:causal-g}, which illustrates the connections 
between rules and their inferred outcomes. 
This structure supports the interpretability of conclusions derived by a symbolic logic solver.
% \begin{equation}
% \begin{aligned}[b]
%     l :& punish \longleftarrow drive, drunk \\
%     m :& punish \longleftarrow resist \\
%     e :& prison \longleftarrow punish \\
%     d :& drive \\
%     k :& drunk \\
%     r :& resist \\
% \end{aligned}
% \label{eq:example-cg}
% \end{equation}
% % https://docs.google.com/drawings/d/1Ing4GFH2kbniiHI5rbXzw-xcAUQAxWAJpgmg4tz1fzc/edit?usp=sharing
% \begin{figure}
%     \centering
%     \includegraphics{assets/causal_g.pdf}
%     \caption{Causal graph of logic program (\cref{eq:example-cg}).}
%     \label{fig:causal-g}
% \end{figure}
\begin{figure}[t]
    \centering
    \begin{minipage}{0.48\textwidth}
    \begin{equation}
    \begin{aligned}[b]
        l :& punish \longleftarrow drive, drunk \\
        m :& punish \longleftarrow resist \\
        e :& prison \longleftarrow punish \\
        d :& drive \\
        k :& drunk \\
        r :& resist \\
    \end{aligned}
    \label{eq:example-cg}
    \end{equation}
    \end{minipage}
    \hfill
    \begin{minipage}{0.45\textwidth}
    \centering
    \includegraphics[width=.5\textwidth]{assets/causal_g.pdf}
    \caption{Causal graph of logic program in \cref{eq:example-cg}.}
    \label{fig:causal-g}
    \end{minipage}
\end{figure}

The tool \src{xclingo}~\cite{Cabalar_2020} 
enables the visualization of justifications for results produced by the \src{clingo} solver. 
It does so by incorporating a trace that corresponds to the rules of the logic program. 
This can help identify potential flaws in \sys's knowledge base and reasoning process.

@@

Evaluation was conducted using a publicly available dataset.\footnote{\url{https://www.kaggle.com/datasets/itachi9604/disease-symptom-description-dataset?select=dataset.csv}}
Each record's list of symptoms serves as input.  
The knowledge base—comprising ASP rules automatically derived  
from medical literature—infers potential diagnoses.  
A prediction is considered correct when the disease inferred  
by the ASP solver matches the actual diagnosis listed in the dataset.

\begin{table}[h]
    \centering
    \caption{Accuracy of the framework on selected diseases. 
    The size column indicates the number of logical terms in the disease-specific ASP program.}
    \begin{tabular}{lcc}
    \toprule
    Diseases & Size & Accuracy \\
    \midrule
    Chickenpox     &  66  & 95\% \\
    Pneumonia      &  75  & 100\% \\
    Common Cold    &  44  & 100\% \\
    \bottomrule
    \end{tabular}
\end{table}

@@

Future work will focus on grounding ASP rules in reliable sources 
through retrieval-augmented generation, fine-tuning LLMs for clinical reasoning, 
and scaling the framework to cover a wider range of diseases.
\newpage

\bibliographystyle{unsrt}
\bibliography{bib}
